A method for defining dispersed community territories

### 2024-06-14 manuscript/jas-reviews-round-two-response.md

### Response to Reviewer 1

1. Dismissiveness in our response.

    Response: We seem to have communicated poorly in our original response as 
    reviewer one had many important and insightful criticisms of our original 
    submission. In fact, we went to great lengths to incorporate their ideas 
    into the current draft. So, we apologize if there was any misunderstanding
    here.

2. Use the 2022 version of Campbell's hiking function.

    Response: It is indeed an unfortunate oversight on our part that we did not 
    use the most up-to-date version of Campbell's hiking function, especially 
    because one of us (Vernon) worked down the hall from him while he was 
    evidently working on the 2022 paper. At any rate, we thank the reviewer for 
    bringing this to our attention and have incorporated it into the updated 
    analysis.

3. Include a figure to clarify how boundaries are drawn.

    Response: Done! This is now included as an additional figure in the 
    manuscript.

### Response to Reviewer 2

We thank the reviewer for their comments and suggestions for this paper.

### Response to Reviewer 4: Reproducibility Review

1. The Github repository is not accessible.

    Response: The repo was initially set to private, but it has now been made
    public.

2. It would be better to include the uncompiled Quarto document alongside the
   compiled HTML.

    Response: This is now done in two ways: The compiled HTML document now
    has the raw Quarto document embedded in it. This can be accessed by hitting 
    the "</> Code" button at the top of the document and selecting "View 
    Source." The raw content can then be copied and pasted into another a
    text file with a .qmd extension and compiled using Quarto. The HTML document
    also now included a link to the Github repository, which also contains the 
    raw Quarto document.

3. Upload all code and data to another trustworthy repository (meaning, not 
   Github).

    Response: We have uploaded all materials to Zenodo. DOI:

4. Include a README that includes all relevant project metadata.

    Response: This is included in the Github repo and is also uploaded to 
    Zenodo.

5. Include the geopackage "community-centers.gpkg."

    Response: This includes sensitive site location data that we do not have the
    authority to make public.

### 2024-06-14 R/community-clustering.qmd
final edits

---
title: "Supplement A: Community Clustering"
date: today
published-title: "COMPILED"
format:
  html: 
    embed-resources: true
    fig-align: center
    fig-dpi: 300
    fig-responsive: true
    number-sections: true
    number-depth: 3
    toc: true
    toc-depth: 3
    smooth-scroll: true
    theme: cosmo
    highlight-style: a11y
execute: 
  echo: true
  warning: false
  error: false
---

@@

  # Campbell's hiking function, see Campbell et al 2019
  # note: campbell function wants degrees!
  campbell <- function(x) {

    # values provided in Supplement Table S3
    # using the 5th percentile of the Lorentz curve as recommended by Campbell
    a <- -1.527
    b <- 14.041
    c <- 36.813
    d <- 0.320
    e <- -0.00273

@@

@@

cmv <- cmv |> define_boundaries(concavity_ratio = 0.1, dilate = c(130, -100))
nrg <- nrg |> define_boundaries(concavity_ratio = 0.1, dilate = c(130, -100))
cib <- cib |> define_boundaries(concavity_ratio = 0.1, dilate = c(130, -100))

>>>>

---
title: "Supplement A: Community Clustering"
date: today
published-title: "COMPILED"
format:
  html: 
    embed-resources: true
    fig-align: center
    fig-dpi: 300
    fig-responsive: true
    number-sections: true
    number-depth: 3
    toc: true
    toc-depth: 3
    smooth-scroll: true
    theme: cosmo
    highlight-style: a11y
    code-tools: true
    code-links:
      - text: kbvernon/community-centers
        icon: github
        href: https://github.com/kbvernon/community-centers
execute: 
  echo: true
  warning: false
  error: false
---

@@

  # Campbell's hiking function, see Campbell et al 2022
  # note: campbell function wants degrees!
  campbell <- function(x) {

    # values provided in Appendix Table S3
    # using the 5th percentile of the Lorentz curve as recommended by Campbell
    a <- -1.4579
    b <- 22.0787
    c <- 76.3271
    d <- 0.0525
    e <- -0.00032002

@@

::: {#fig-boundary-example}
![](../figures/boundary-example.png){width="100%"
fig-alt="An example of the method used to define community boundaries."}

Example of the method used to define community boundaries. Left panel shows two
hull geometries: (i) convex hull with dark gray fill and (ii) concave hull 
(ratio = 0.1) with light gray fill. Middle panel shows every pairwise short path 
between the points along the outer edge of the convex hull. Right panel shows 
community boundary derived by applying the concave hull (ratio = 0.5) to the 
combined set of site points and vertices of short paths. The most notable
differences are along the left boundary.
:::

@@

cmv <- cmv |> define_boundaries(concavity_ratio = 0.5, dilate = c(130, -100))
nrg <- nrg |> define_boundaries(concavity_ratio = 0.5, dilate = c(130, -100))
cib <- cib |> define_boundaries(concavity_ratio = 0.5, dilate = c(130, -100))

### 2024-06-14 manuscript/manuscript.qmd
final edits

So far, we have only grouped site points into clusters or communities. The goal,
however, is to define their spatial extent. That means, for each set of site
points, we need some way of drawing a polygon that encompasses all of them. This
can be done in a number of ways. The simplest strategy would be to find the
centroid of the set of points and draw a circle with the smallest radius that
still includes them all. This would be utterly arbitrary, however, and frankly
unrealistic, not representing any sort of meaningful boundary given the
landscape. An alternative would be to use the convex hull of the set of points,
but this tends to exaggerate the total area of the community, especially when
the distribution of points is suggestive of a concave shape, like the letter
'C'. A concave hull would handle peculiar shapes, but it would also introduce
the opposite problem, restricting the area of the community to an unreasonable
degree. Our somewhat brute force way of balancing these factors is to identify
paths with shortest commute times between the outer most farms in a community,
extract the vertices from those paths, and then incorporate those into the set
of points used to define the concave hull. The result is a polygon that is
concave as well in the dimension of time, meaning the paths with shortest
commute times between any two farms never leave the community.

@@

The key to our path analysis is to take information about the terrain in our
study regions and convert it first into meaningful estimates of hiking speed and
then into meaningful estimates of travel time. Fortunately, attempts to model
hiking speed across a range of slopes have advanced considerably in recent
years, most notably in @campbell_etal2019. Drawing on a large, crowd-sourced
fitness tracker data set consisting of approximately 420,000 individual hikes,
jogs, and trail runs from nearly 30,000 individuals in Salt Lake City, UT,
@campbell_etal2019 offer what is easily the most empirically robust estimate of
hiking speed to date. Our application of Campbell's hiking function relies on
coefficients from the fifth percentile of individuals in their sample, using the
Cauchy (or Lorentz) distribution as recommended by Campbell, and assumes that
everyone in the areas of our analysis were largely equal in their walking speed
and endurance. This makes their walking speeds comparable to, but not the same
as, what would be estimated by Tobler's hiking function [@tobler1993].

@@

The goals of this analysis are (i) to identify dispersed farming communities
based on the distribution of farmsteads and their proximity to community centers
and then (ii) to regionalize those communities by giving them meaningful
boundaries. This was done by developing a simple algorithm inspired by the
concept of a CBSA combined with previous work by @varien1999 and
@reese_etal2019. Using parameter values $D{\text -}join$ = 30 minutes,
$D{\text -}max$ = 1 hour, and $P$ = 0.8, our algorithm identified 104
communities in CMV, 90 communities in NRG, and 45 communities in CIB (see
@fig-communities). The number of communities removed because they included an
insufficient number of sites was 4 for CMV, 35 for NRG, and 23 for CIB, leaving
100, 55, and 22 communities with definable spatial extents.

The mean areas of these communities are 8.68 km$^2$ in CMV, 6.06 km$^2$ in NRG,
and 7.19 km$^2$ in CIB; the mean room counts (for all community centers and
farms) are 647.76 in CMV, 593.49 in NRG, and 766.68 in CIB; and the mean room
densities are 99.69 rooms per km$^2$ in CMV, 219.77 rooms per km$^2$ in NRG, and
238.76 rooms per km$^2$ in CIB. We emphasize that the means in several cases
deviate to a considerable degree from median values (denoted in @tbl-results
with the Greek letter $\eta$), indicating skew in the data and the potential
presence of large outliers. This is illustrated in the boxplots in
@fig-densities, which show the distributions of these and other variables of
interest that can be derived from our analysis and also used to evaluate its
merits (see also @tbl-results). Note that these summaries are time-insensitive.
They represent the total archaeological sample in each region.

@@

\begin{singlespace}
\small
\begin{tabular}{lrllllll}
    & 
    & 
\textbf{\begin{tabular}[c]{@{}l@{}}Area\\ (km2)\end{tabular}} & \textbf{\begin{tabular}[c]{@{}l@{}}Rooms\\ (N)\end{tabular}} & \textbf{\begin{tabular}[c]{@{}l@{}}Farms\\ (N)\end{tabular}} & \textbf{\begin{tabular}[c]{@{}l@{}}Centers\\ (N)\end{tabular}} & \textbf{\begin{tabular}[c]{@{}l@{}}Commute\\ (mins)\end{tabular}}  & \textbf{\begin{tabular}[c]{@{}l@{}}Room Density\\ (N/km2)\end{tabular}} 
  \\[2mm]
  \hline
  \noalign{\vskip2mm}
CMV &
\begin{tabular}[c]{@{}l@{}} $\eta$\\ $\mu$\\ $\sigma$ \end{tabular} &
\begin{tabular}[c]{@{}l@{}} : 7.1\\ : 8.68\\ : 6.73 \end{tabular} &
\begin{tabular}[c]{@{}l@{}} : 402\\ : 647.76\\ : 701.36 \end{tabular} &
\begin{tabular}[c]{@{}l@{}} : 24.5\\ : 44.38\\ : 55.86 \end{tabular} &
\begin{tabular}[c]{@{}l@{}} : 1\\ : 1.65\\ : 1.37 \end{tabular} &
\begin{tabular}[c]{@{}l@{}} : 22.35\\ : 23.28\\ : 8.83 \end{tabular} &
\begin{tabular}[c]{@{}l@{}} : 57.49\\ : 99.69\\ : 97.36 \end{tabular}
  \\[5mm]
  \hline
  \noalign{\vskip2mm}
NRG &
\begin{tabular}[c]{@{}l@{}} $\eta$\\ $\mu$\\ $\sigma$ \end{tabular} &
\begin{tabular}[c]{@{}l@{}} : 5.53\\ : 6.06\\ : 4.56 \end{tabular} &
\begin{tabular}[c]{@{}l@{}} : 479\\ : 593.49\\ : 444.7 \end{tabular} &
\begin{tabular}[c]{@{}l@{}} : 20\\ : 36.73\\ : 34.96 \end{tabular} &
\begin{tabular}[c]{@{}l@{}} : 2\\ : 2.49\\ : 2.06 \end{tabular} &
\begin{tabular}[c]{@{}l@{}} : 19.12\\ : 19.04\\ : 8.77 \end{tabular} &
\begin{tabular}[c]{@{}l@{}} : 99.43\\ : 219.77\\ : 450.86 \end{tabular}
  \\[5mm]
  \hline
  \noalign{\vskip2mm}
CIB &
\begin{tabular}[c]{@{}l@{}} $\eta$\\ $\mu$\\ $\sigma$ \end{tabular} &
\begin{tabular}[c]{@{}l@{}} : 4.37\\ : 7.19\\ : 7.39 \end{tabular} &
\begin{tabular}[c]{@{}l@{}} : 492\\ : 766.68\\ : 626.09 \end{tabular} &
\begin{tabular}[c]{@{}l@{}} : 13.5\\ : 24.14\\ : 25.34 \end{tabular} &
\begin{tabular}[c]{@{}l@{}} : 2\\ : 2\\ : 1.11 \end{tabular} &
\begin{tabular}[c]{@{}l@{}} : 21.16\\ : 24.74\\ : 12.27 \end{tabular} &
\begin{tabular}[c]{@{}l@{}} : 170.25\\ : 238.76\\ : 270.64 \end{tabular}
  \\[5mm]
  \hline
  \noalign{\vskip2mm}
\end{tabular}
\end{singlespace}
```
$\eta$: median, $\mu$: mean, $\sigma$: standard deviation

@@

Results of the sensitivity analysis are as expected given the role that the main
tuning parameters play in the algorithm (see @fig-sensitivity). Increasing
$D{\text -}join$ makes it easier to join two communities, thus leading to growth
in the total area of communities and declines in the total number of resulting
communities. Increasing $P$ makes it harder to join communities, thus leading to
fewer and smaller communities. Increasing $D{\text -}max$ increases the total
area of communities and the average commute time to centers as it serves to
filter out distant farmsteads. Importantly, finer-grained grid resolutions had
no meaningful impact on the results of this analysis, so the level of
aggregation we chose appears acceptable. One puzzling result is that the CIB
region shows some volatility in its room densities with respect to the main
tuning parameters. The number of CIB communities is also largely invariant. We
comment on this in the discussion.

@@

Figure shows the difference between the log density of commute times (in
minutes) from observed farms and random locations to the nearest community
center for each region, along with the location of their 95% quantile (for
observed farms, not random points).
:::

>>>>

So far, we have only grouped site points into clusters or communities. The goal,
however, is to define their spatial extent. That means, for each set of site
points, we need some way of drawing a polygon that encompasses all of them. This
can be done in a number of ways. The simplest strategy would be to find the
centroid of the set of points and draw a circle with the smallest radius that
still includes them all. This would be utterly arbitrary, however, and frankly
unrealistic, not representing any sort of meaningful boundary given the
landscape. An alternative would be to use the convex hull of the set of points,
but this tends to exaggerate the total area of the community, especially when
the distribution of points is suggestive of a concave shape, like the letter
'C'. A concave hull would handle peculiar shapes, but it could also introduce
the opposite problem, restricting the area of the community to an unreasonable
degree, depending on how strict the convexity ratio is. Our somewhat brute force
way of balancing these factors is to identify paths with shortest commute times
between the outer most farms in a community, extract the vertices from those
paths, and then incorporate those into the set of points used to define the
concave hull (see @fig-boundary-example). The result is a polygon that is
concave as well in the dimension of time, meaning the paths with shortest
commute times between any two farms never leave the community.

::: {#fig-boundary-example}
![](../figures/boundary-example.png){width="6.5in"}

Example of the method used to define community boundaries. Left panel shows two
hull geometries: (i) convex hull with dark gray fill and (ii) concave hull
(ratio = 0.1) with light gray fill. Middle panel shows every pairwise short path
between the points along the outer edge of the convex hull. Right panel shows
community boundary derived by applying the concave hull (ratio = 0.5) to the
combined set of site points and vertices of short paths. The most notable
differences are along the left boundary.
:::

@@

The key to our path analysis is to take information about the terrain in our
study regions and convert it first into meaningful estimates of hiking speed and
then into meaningful estimates of travel time. Fortunately, attempts to model
hiking speed across a range of slopes have advanced considerably in recent
years, most notably in [@campbell_etal2022; see also @campbell_etal2019]. Our
application of Campbell's hiking function relies on coefficients from the median
range of individuals in their sample and assumes that everyone in the areas of
our analysis was largely equal in their walking speed and endurance. This makes
their walking speeds comparable to, but not the same as, what would be estimated
by Tobler's hiking function [@tobler1993].

@@

The goals of this analysis are (i) to identify dispersed farming communities
based on the distribution of farmsteads and their proximity to community centers
and then (ii) to regionalize those communities by giving them meaningful
boundaries. This was done by developing a simple algorithm inspired by the
concept of a CBSA combined with previous work by @varien1999 and
@reese_etal2019. Using parameter values $D{\text -}join$ = 30 minutes,
$D{\text -}max$ = 1 hour, and $P$ = 0.8, our algorithm identified 103
communities in CMV, 88 communities in NRG, and 45 communities in CIB (see
@fig-communities). The number of communities removed because they included an
insufficient number of sites was 5 for CMV, 33 for NRG, and 23 for CIB, leaving
98, 55, and 22 communities with definable spatial extents.

The mean areas of these communities are 9.51 km$^2$ in CMV, 6.5 km$^2$ in NRG,
and 7.74 km$^2$ in CIB; the mean room counts (for all community centers and
farms) are 663 in CMV, 598.65 in NRG, and 768.05 in CIB; and the mean room
densities are 83.83 rooms per km$^2$ in CMV, 183.93 rooms per km$^2$ in NRG, and
207.18 rooms per km$^2$ in CIB. We emphasize that the means in several cases
deviate to a considerable degree from median values (denoted in @tbl-results
with the Greek letter $\eta$), indicating skew in the data and the potential
presence of large outliers. This is illustrated in the boxplots in
@fig-densities, which show the distributions of these and other variables of
interest that can be derived from our analysis and also used to evaluate its
merits (see also @tbl-results). Note that these summaries are time-insensitive.
They represent the total archaeological sample in each region.

@@

\begin{singlespace}
\small
\begin{tabular}{lrllllll}
    & 
    & 
\textbf{\begin{tabular}[c]{@{}l@{}}Area\\ (sq.km)\end{tabular}} & \textbf{\begin{tabular}[c]{@{}l@{}}Rooms\\ (N)\end{tabular}} & \textbf{\begin{tabular}[c]{@{}l@{}}Farms\\ (N)\end{tabular}} & \textbf{\begin{tabular}[c]{@{}l@{}}Centers\\ (N)\end{tabular}} & \textbf{\begin{tabular}[c]{@{}l@{}}Commute\\ (mins)\end{tabular}}  & \textbf{\begin{tabular}[c]{@{}l@{}}Room Density\\ (N/sq.km)\end{tabular}} 
  \\[2mm]
  \hline
  \noalign{\vskip2mm}
CMV &
\begin{tabular}[c]{@{}l@{}} $\eta$\\ $\mu$\\ $\sigma$ \end{tabular} &
\begin{tabular}[c]{@{}l@{}} : 7.61\\ : 9.51\\ : 7.2 \end{tabular} &
\begin{tabular}[c]{@{}l@{}} : 399\\ : 663\\ : 753.62 \end{tabular} &
\begin{tabular}[c]{@{}l@{}} : 25.5\\ : 45.61\\ : 59.11 \end{tabular} &
\begin{tabular}[c]{@{}l@{}} : 1\\ : 1.67\\ : 1.39 \end{tabular} &
\begin{tabular}[c]{@{}l@{}} : 21.51\\ : 22.72\\ : 8.3 \end{tabular} &
\begin{tabular}[c]{@{}l@{}} : 55.08\\ : 83.83\\ : 71.93 \end{tabular}
  \\[5mm]
  \hline
  \noalign{\vskip2mm}
NRG &
\begin{tabular}[c]{@{}l@{}} $\eta$\\ $\mu$\\ $\sigma$ \end{tabular} &
\begin{tabular}[c]{@{}l@{}} : 5.45\\ : 6.5\\ : 5.07 \end{tabular} &
\begin{tabular}[c]{@{}l@{}} : 475\\ : 598.65\\ : 468.12 \end{tabular} &
\begin{tabular}[c]{@{}l@{}} : 18\\ : 36.91\\ : 37.55 \end{tabular} &
\begin{tabular}[c]{@{}l@{}} : 2\\ : 2.53\\ : 2.17 \end{tabular} &
\begin{tabular}[c]{@{}l@{}} : 18.14\\ : 18.28\\ : 8.65 \end{tabular} &
\begin{tabular}[c]{@{}l@{}} : 98.69\\ : 183.93\\ : 269.84 \end{tabular}
  \\[5mm]
  \hline
  \noalign{\vskip2mm}
CIB &
\begin{tabular}[c]{@{}l@{}} $\eta$\\ $\mu$\\ $\sigma$ \end{tabular} &
\begin{tabular}[c]{@{}l@{}} : 4.59\\ : 7.74\\ : 7.96 \end{tabular} &
\begin{tabular}[c]{@{}l@{}} : 492\\ : 768.05\\ : 626.91 \end{tabular} &
\begin{tabular}[c]{@{}l@{}} : 13.5\\ : 24.23\\ : 25.39 \end{tabular} &
\begin{tabular}[c]{@{}l@{}} : 2\\ : 2\\ : 1.11 \end{tabular} &
\begin{tabular}[c]{@{}l@{}} : 20.85\\ : 24.36\\ : 12.14 \end{tabular} &
\begin{tabular}[c]{@{}l@{}} : 157.01\\ : 207.18\\ : 218.53 \end{tabular}
  \\[5mm]
  \hline
  \noalign{\vskip2mm}
\end{tabular}
\end{singlespace}
```
$\eta$: median, $\mu$: mean, $\sigma$: standard deviation

@@

Results of the sensitivity analysis are as expected given the role that the main
tuning parameters play in the algorithm (see @fig-sensitivity). Increasing
$D{\text -}join$ makes it easier to join two communities, thus leading to growth
in the total area of communities and declines in the total number of resulting
communities. Increasing $P$ makes it harder to join communities, thus leading to
more smaller communities. Increasing $D{\text -}max$ increases the total area of
communities and the average commute time to centers as it serves to filter out
distant farmsteads. Importantly, finer-grained grid resolutions had no
meaningful impact on the results of this analysis, so the level of aggregation
we chose appears acceptable. One puzzling result is that the CIB region shows
some volatility in its room densities with respect to the main tuning
parameters. The number of CIB communities is also largely invariant. We comment
on this in the discussion.

@@

Figure shows the difference between the log density of commute times (in
minutes) from observed farms and random locations to the nearest community
center for each region, along with the location of their 95% quantile for
observed farms (the location of the point next to the respective region label,
all being between 60 and 90 minutes).
:::
